---
title: "pre-post designs"
author: "Jeffrey A. Walker"
date: "11/13/2020"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    fig_caption: yes
    number_sections: true
    code_folding: show
    includes:
      before_body: copyright.html
---

# Setup and Import

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(here)
library(janitor)
library(readxl)
library(data.table)

# graphics and tables
library(ggplot2)
library(ggpubr)
library(ggsci)
library(ggthemes)
library(cowplot)
library(kableExtra)

# analysis packages
library(mvtnorm)
library(nlme)
library(lmerTest)
library(emmeans)
library(afex)

pal_okabe_ito <- colorblind_pal()(8)[2:8] # ggthemes
pal_okabe_ito_blue <- pal_okabe_ito[c(5,6,1,2,3,7,4)] 
pal_okabe_ito_red <- pal_okabe_ito[c(6,5,3,1,2,7,4)] 

here <- here::here
data_folder <- "content/data"
output_folder <- "content/output"

```

# simulation functions
## pre-post fake data generator 

```{r pre-post-FDG}

generate_pre_post <- function(
  n = 6, # if vector then sample per group
  sigma_0 = 1, # error at time 0
  sigma_1 = 1, # error at time 1
  rho = 0.6, # correlation between time 0 and 1
  beta_0 = 10, # control mean at time 0
  beta_1 = 0, # treatment effect at time 0
  beta_2 = 5, # time effect at time 1
  beta_3 = 1 # interaction effect -- this is the treatment effect
){
  
  if(length(n) == 1){n <- rep(n,2)} # set n = n for all groups
  n_list <- n
  N <- sum(n_list)
  
  # correlated error matrix
  R <- matrix(c(1, rho, rho, 1), nrow=2)
  L = matrix(c(sigma_0, 0, 0, sigma_1), nrow=2)
  Sigma <- L%*%R%*%L
  e <- rmvnorm(N, mean = c(0,0), sigma = Sigma) %>%
    data.table()
  e[, id := paste0("id_", 1:N)]
  e_long <- melt(e,
                 id.vars = "id")
    
  fd <- data.table(
    treatment = factor(rep(rep(c("cn", "tr"), n_list), 2)),
    time = factor(rep(c("pre", "post"), each = N),
                  levels = c("pre", "post"))
  )
  
  # fd is in long format
  X <- model.matrix(~treatment*time, data=fd)
  beta <- c(beta_0, beta_1, beta_2, beta_3)
  fd[, y := (X%*%beta)[,1] + e_long[, value]]
  fd[, id := e_long[, id]]
  fd[, id := factor(id)]
 
  return(fd)
}

```

## test FDG

```{r test-pre-post-FDG, eval=FALSE}
fd <- generate_pre_post(
  n = c(10^5, 10^5),
  sigma_0 = 1,
  sigma_1 = 2,
  rho = 0.6, 
  beta_0 = 10, 
  beta_1 = 1,
  beta_2 = 2,
  beta_3 = 1
)

m1 <- lm(post ~ treatment, data = fd)
fd[, post_cond := coef(m1)[1] + residuals(m1)]
coef(summary(m1))
cov(fd[, .SD, .SDcols=c("pre", "post_cond")])
cor(fd[, .SD, .SDcols=c("pre", "post_cond")])
coef(summary(lm(post ~ pre + treatment, data=fd)))

fd_long <- melt(fd,
                id.vars = c("treatment", "id"),
                measure.vars = c("pre", "post"),
                variable.name = "time",
                value.name = "y")
fd_long[, time := factor(time, c("pre", "post"))]
coef(summary(lm(y ~ treatment*time, data = fd_long)))

```



# Fit model functions
## change-score

```{r lm-change}
lm_change <- function(dt){
  fit <- lm(change ~ treatment, data = dt)
  return(fit)
}
```

## two-way anova

```{r two-way anova}
two_way <- function(dt_long){
  fit <- lm(y ~ treatment*time, data=dt_long)
  return(fit)
}
```

## repeated measures anova

```{r repeated-measures-anova, message=FALSE}
rmanova <- function(dt_long){
  fit <- aov_4(y ~ treatment*time + (time|id),
               data=dt_long)
  return(fit)
}
```

## linear model with baseline as covariate

```{r lm-cov}
lm_cov <- function(dt){
  fit <- lm(post ~ pre + treatment, data = dt)
  return(fit)
}
```

## linear model with correlated error

```{r lda}
lda <- function(dt_long){
  fit <- gls(y ~ treatment*time,
             data = dt_long,
             weights = varIdent(form = ~ 1 | time),
             correlation= corSymm(form =~ 1 | id))
  return(fit)
}
```


```{r lda-random}
lda_random <- function(dt_long){
  fit <- lme(y ~ treatment*time,
             random = ~1|id,
             data = dt_long,
             weights = varIdent(form= ~ 1 | time),
             correlation= corSymm(form=~ 1 | id))
  return(fit)
}
```

```{r lmm}
lmm <- function(dt_long){
  fit <- lmer(y ~ treatment*time + (1|id),
             data = dt_long)
  return(fit)
}
```

note lmm, rmanova, and lda_random are effectively the same, at least for my parameter space

# simulations
## single post time point ("pre-post")

```{r simulation-single-post-time-point, message=FALSE}

simulate_pre_post <- function(
  niter = 1000, # number of iterations
  n = 6, # if vector then sample per group
  sigma_0 = 1, # error at time 0
  sigma_1 = 1, # error at time 1
  rho = 0.6, # correlation between time 0 and 1
  beta_0 = 10, # control mean at time 0
  beta_1 = 1,
  beta_2 = 5, # time effect at time 1
  beta_3 = 1  # interaction
){
  
  model_list <- c("lm_cov", "lm_change", "lda", "rmanova")
  model_params <- c("b1", "b3", "lower", "upper", "p")
  res_cols <- paste(rep(model_list, each = length(model_params)),
                    model_params,
                    sep = ".")
  prob_wide <- data.table(matrix(as.numeric(NA),
                                 nrow=niter,
                                 ncol=length(res_cols)))
  colnames(prob_wide) <- res_cols
  
  if(length(n) == 1){n <- rep(n,2)} # set n = n for all groups
  n_list <- n
  N <- sum(n_list)
  
  # correlated error matrix
  R <- matrix(c(1, rho, rho, 1), nrow=2)
  L = matrix(c(sigma_0, 0, 0, sigma_1), nrow=2)
  Sigma <- L%*%R%*%L

  fd_long <- data.table(
    treatment = factor(rep(rep(c("cn", "tr"), n_list), 2)),
    time = factor(rep(c("pre", "post"), each = N),
                  levels = c("pre", "post"))
  )
  
  # fd is in long format
  X <- model.matrix(~ treatment*time, data = fd_long)
  beta <- c(beta_0, beta_1, beta_2, beta_3)
  
  for(iter in 1:niter){
    # fd_long <- generate_pre_post(
    #   n = n,
    #   sigma_0 = sigma_0,
    #   sigma_1 = sigma_1,
    #   rho = rho, 
    #   beta_0 = beta_0, 
    #   beta_1 = beta_1,
    #   beta_2 = beta_2,
    #   beta_3 = beta_3
    # )
    
    e <- rmvnorm(N, mean = c(0,0), sigma = Sigma) %>%
      data.table()
    e[, id := paste0("id_", 1:N)]
    e_long <- melt(e,
                   id.vars = "id")
    
    fd_long[, y := (X%*%beta)[,1] + e_long[, value]]
    fd_long[, id := e_long[, id]]
    fd_long[, id := factor(id)]
    
    
    # reshape to wide
    fd_wide <- dcast(fd_long,
                treatment + id ~ time,
                value.var = "y")
    fd_wide[, change := post - pre]
    
    if("lm_cov" %in% model_list){
      m1 <- lm_cov(fd_wide)
      coef_table <- coef(summary(m1))
      ci_table <- confint(m1)
      prob_wide[iter, lm_cov.b3 := coef_table["treatmenttr", "Estimate"]]
      prob_wide[iter, lm_cov.lower := ci_table["treatmenttr", "2.5 %"]]
      prob_wide[iter, lm_cov.upper := ci_table["treatmenttr", "97.5 %"]]
      prob_wide[iter, lm_cov.p := coef_table["treatmenttr", "Pr(>|t|)"]]
    }
    
   if("lm_change" %in% model_list){
    m1 <- lm_change(fd_wide)
    coef_table <- coef(summary(m1))
    ci_table <- confint(m1)
    prob_wide[iter, lm_change.b3 := coef_table["treatmenttr", "Estimate"]]
    prob_wide[iter, lm_change.lower := ci_table["treatmenttr", "2.5 %"]]
    prob_wide[iter, lm_change.upper := ci_table["treatmenttr", "97.5 %"]]
    prob_wide[iter, lm_change.p := coef_table["treatmenttr", "Pr(>|t|)"]]
   }
    
   if("lda" %in% model_list){
      #m2 <- lda(fd_long)
      m2 <- gls(y ~ treatment*time,
                data = fd_long,
                weights = varIdent(form = ~ 1 | time),
                correlation= corSymm(form = ~ 1 | id))
      coef_table <- coef(summary(m2))
      # ci_table <- confint(m2)
      skip_to_next <- FALSE
      m2_emm <- tryCatch(emmeans(m2,
                                 specs = c("treatment", "time"),
                                 data = fd_long,
                                 mode = "satterthwaite"),
                         error = function(e) { skip_to_next <- TRUE})
      if(skip_to_next == TRUE){ next }
      m2_pairs <- contrast(m2_emm, interaction = c("consec"), by = NULL) %>%
        summary(infer = TRUE) %>%
        data.table
      prob_wide[iter, lda.b1 := coef_table["treatmenttr", "Value"]]
      prob_wide[iter, lda.b3 := m2_pairs[1, estimate]]
      prob_wide[iter, lda.lower := m2_pairs[1, lower.CL]]
      prob_wide[iter, lda.upper := m2_pairs[1, upper.CL]]
      prob_wide[iter, lda.p := m2_pairs[1, p.value]]
    }
    
    if("lda_r" %in% model_list){
      m3 <- lda_random(fd_long)
      prob_wide[iter, lda_r.p := coef(summary(m3))["treatmenttr:timepost", "p-value"]]
    }
    
    if("lmm" %in% model_list){
      m4 <- lmm(fd_long)
      prob_wide[iter, lmm.p := coef(summary(m4))["treatmenttr:timepost", "Pr(>|t|)"]]
    }
    
    if("rmanova" %in% model_list){
      m5 <- rmanova(fd_long)
      prob_wide[iter, rmanova.p := summary(m5)$univariate.tests["treatment:time", "Pr(>F)"]]
    }
  }

  return(prob_wide)
}
```

## Summary statistics

```{r summary stats}
p_summary <- function(prob, niter, by_list){
  prob_summary <- prob[, .(freq_lt_05 = sum(p_value < 0.05)/max(niter)),
          by = .(method)]
  return(prob_summary)
}

```

## Plot: 

```{r plot-it}
plot_it <- function(prob){
  jco_pal <- pal_jco()(6)
  group_colors <- jco_pal[c(2,5,4,6)]
  group_shapes <- c(15, 16, 17,  18)
  
  pd <- position_dodge(0.8)
  gg1 <- ggplot(data = prob,
                aes(x=method, y=freq_lt_05, color=method, shape=method)) +
    geom_point(position = pd) +
    scale_color_manual(values = group_colors) +
    scale_shape_manual(values = group_shapes) +
    ylab("Frequency p < 0.05") +
    xlab("") +
    theme_pubr() +
    theme(legend.position="none") +
    guides(col = guide_legend(ncol = 5, byrow = FALSE)) +
    NULL
  return(gg1)
}
```

## Type I error

```{r type I, message=FALSE, eval=TRUE}
do_it <- TRUE
set.seed(1)
niter <- 1000 # number of iterations

if(do_it == TRUE){

# parameters of simulation
  n1_list <- c(6, 30) # sample size
  n2_list <- as.numeric(NA) # fraction more than n1
  sigma_0 <- 1 # error at time 0
  sigma_1_list <- c(1, 2) # error at time 1
  rho_list <- c(0.3, 0.8) # correlation between time 0 and 1
  beta_0 <- 10 # control mean at time 0
  beta_1_list <- c(0.5, 1)
  beta_2 <- 5 # time effect at time 1
  beta_3_list <- c(0) # time effect at time 1
 
  param_combis <- expand.grid(n1 = n1_list,
                              n2 = n2_list,
                              sigma_1 = sigma_1_list,
                              rho = rho_list,
                              beta_1 = beta_1_list,
                              beta_3 = beta_3_list) %>%
    data.table()
  prob_type1_table <- data.table(NULL)
  for(sim_i in 1:nrow(param_combis)){
    n1_i <- param_combis[sim_i, n1]
    # n2_i <- n1_i + round(param_combis[sim_i, n2]*n1_i, 0)
    n2_i <- n1_i
    sigma_1_i <- param_combis[sim_i, sigma_1]
    rho_i <- param_combis[sim_i, rho]
    beta_1_i <- param_combis[sim_i, beta_1]
    beta_3_i <- param_combis[sim_i, beta_3]
    prob_type1 <- simulate_pre_post(niter = niter,
                                    n = c(n1_i, n2_i),
                                    sigma_0 = sigma_0,
                                    sigma_1 = sigma_1_i,
                                    rho = rho_i,
                                    beta_0 = beta_0,
                                    beta_1 = beta_1_i,
                                    beta_2 = beta_2,
                                    beta_3 = beta_3_i)
    prob_type1_table <- rbind(prob_type1_table,
                              data.table(n1 = n1_i,
                                         n2 = n2_i,
                                         sigma_1 = sigma_1_i,
                                         rho = rho_i,
                                         beta_1 = beta_1_i,
                                         prob_type1))
  }
  outfile_name <- "prob_type1_table.rds"
  out_path <- here(output_folder, outfile_name)
  saveRDS(prob_type1_table, out_path)
}else{
  outfile_name <- "prob_type1_table.rds"
  out_path <- here(output_folder, outfile_name)
  prob_type1_table <- readRDS(out_path)
}

```

## Power

```{r Power sim, message=FALSE, eval=TRUE}
do_it <- TRUE
set.seed(2)
niter <- 1000 # number of iterations

if(do_it == TRUE){
  
# parameters of simulation
  n1_list <- c(6, 30) # sample size
  n2_list <- c(0) # fraction more than n1
  sigma_0 <- 1 # error at time 0
  sigma_1_list <- c(2) # error at time 1
  rho_list <- c(0.3, 0.8) # correlation between time 0 and 1
  beta_0 <- 10 # control mean at time 0
  beta_1_list <- c(0.5)
  beta_2 <- 5 # time effect at time 1 
  beta_3_list <- c(.5, 1) # time effect at time 1
  
  param_combis <- expand.grid(n1 = n1_list,
                              n2 = n2_list,
                              sigma_1 = sigma_1_list,
                              rho = rho_list,
                              beta_1 = beta_1_list,
                              beta_3 = beta_3_list) %>%
    data.table()
  prob_power_table <- data.table(NULL)
  for(sim_i in 1:nrow(param_combis)){
    n1_i <- param_combis[sim_i, n1]
    # n2_i <- n1_i + round(param_combis[sim_i, n2]*n1_i, 0)
    n2_i <- n1_i
    sigma_1_i <- param_combis[sim_i, sigma_1]
    rho_i <- param_combis[sim_i, rho]
    beta_1_i <- param_combis[sim_i, beta_1]
    beta_3_i <- param_combis[sim_i, beta_3]
    prob_power <- simulate_pre_post(niter = niter,
                                    n = c(n1_i, n2_i),
                                    sigma_0 = sigma_0,
                                    sigma_1 = sigma_1_i,
                                    rho = rho_i,
                                    beta_0 = beta_0,
                                    beta_1 = beta_1_i,
                                    beta_2 = beta_2,
                                    beta_3 = beta_3_i)
    prob_power_table <- rbind(prob_power_table,
                              data.table(n2 = n2_i,
                                         sigma_1 = sigma_1_i,
                                         rho = rho_i,
                                         beta_1 = beta_1_i,
                                         beta_3 = beta_3_i,
                                         prob_power))
  }
  outfile_name <- "prob_power_table.rds"
  out_path <- here(output_folder, outfile_name)
  saveRDS(prob_power_table, out_path)
}else{
  outfile_name <- "prob_power_table.rds"
  out_path <- here(output_folder, outfile_name)
  prob_power_table <- readRDS(out_path)
}

```

# Plots
## Type I

```{r}
# p-values
prob_type1_p <- melt(prob_type1_table,
                     id.vars = c("n2", "sigma_1", "beta_1", "rho"),
                     measure.vars = c("lm_cov.p", "lm_change.p", "lda.p", "rmanova.p"),
                     variable.name = "model",
                     value.name = "p_value")
prob_type1_table <- prob_type1_p[,
                               .(freq_lt_05 = sum(p_value < 0.05)/niter),
                               by = .(model, n2, sigma_1, beta_1, rho)]
prob_type1_table[, sim := factor(rep(1:16, 4))]

```

```{r prob_type1_table}
pal_okabe_ito_man <- pal_okabe_ito[c(5,7,3,6,1,2,4)]

pd <- position_dodge(0.2)
base <- 0.067
spacer <- 0.002
row <- 0
ggplot(data = prob_type1_table[model != "lm_cov.p"],
       aes(x = sim,
           y = freq_lt_05,
           color = model)) +
  geom_point(size = 3, position = pd) +
  geom_line(aes(group = model), position = pd) +
  
  annotate(geom = "text",
           label = c("n:"),
           x = 16.5,
           y = base + (row+0)*spacer
  ) +
  annotate(geom = "text",
           label = prob_type1_table[1:16, n2],
           x = 1:16,
           y = base + (row+0)*spacer
  ) +
  
  annotate(geom = "text",
           label = c("sigma_1:"),
           x = 16.5,
           y = base + (row+1)*spacer
  ) +
  annotate(geom = "text",
           label = prob_type1_table[1:16, sigma_1],
           x = 1:16,
           y = base + (row+1)*spacer
  ) +
  
  annotate(geom = "text",
           label = c("rho:"),
           x = 16.5,
           y = base + (row+2)*spacer
  ) +
  annotate(geom = "text",
           label = prob_type1_table[1:16, rho],
           x = 1:16,
           y = base + (row+2)*spacer
  ) +
 
  annotate(geom = "text",
           label = c("beta_1:"),
           x = 16.5,
           y = base + (row+3)*spacer
  ) +
  annotate(geom = "text",
           label = prob_type1_table[1:16, beta_1],
           x = 1:16,
           y = base + (row+3)*spacer
  ) +

 
  #coord_cartesian(xlim(c(0, 9))) +
  scale_color_manual(values = pal_okabe_ito_man,
                     name = NULL) +

  theme_pubr() +
  NULL

```

## Estimates
```{r estimates-table}
# estimates
estimate_b3 <- melt(prob_power_table,
                id.vars = c("n2", "sigma_1", "beta_1", "beta_3", "rho"),
                measure.vars = list(c("lm_cov.b3", "lm_change.b3", "lda.b3"),
                                    c("lm_cov.lower", "lm_change.lower", "lda.lower"),
                                    c("lm_cov.upper", "lm_change.upper", "lda.upper")),
                variable.name = "model",
                value.name = c("b3", "lower", "upper"))
estimate_b3[model == "1", model := "lm_cov"]
estimate_b3[model == "2", model := "lm_change"]
estimate_b3[model == "3", model := "lda"]

estimate_table <- estimate_b3[, .(mean = mean(b3),
                               lower = quantile(b3, 0.025),
                               upper = quantile(b3, 0.975)),
                               by = .(n2, sigma_1, rho, beta_1, beta_3, model)]
estimate_table[, sim := factor(rep(1:8, 3))]

```

```{r estimates-plot}
pal_okabe_ito_man <- pal_okabe_ito[c(5,7,3,6,1,2,4)]

pd <- position_dodge(0.2)
base <- 1.3
spacer <- 0.075
ggplot(data = estimate_table,
       aes(x = sim,
           y = mean,
           color = model)) +
  geom_point(size = 3, position = pd) +
  geom_line(aes(group = model), position = pd) +
  
  annotate(geom = "text",
           label = c("n:"),
           x = 8.5,
           y = base + 2*spacer
  ) +
  annotate(geom = "text",
           label = estimate_table[1:8, n2],
           x = 1:8,
           y = base + 2*spacer
  ) +
  
  annotate(geom = "text",
           label = c("rho:"),
           x = 8.5,
           y = base + 1*spacer
  ) +
  annotate(geom = "text",
           label = estimate_table[1:8, rho],
           x = 1:8,
           y = base + 1*spacer
  ) +
 
  annotate(geom = "text",
           label = c("beta_3:"),
           x = 8.5,
           y = base + 0*spacer
  ) +
  annotate(geom = "text",
           label = estimate_table[1:8, beta_3],
           x = 1:8,
           y = base + 0*spacer
  ) +
 
  #coord_cartesian(xlim(c(0, 9))) +
  scale_color_manual(values = pal_okabe_ito_man,
                     name = NULL) +

  theme_pubr() +
  NULL

```

## Coverage

```{r}
# coverage -- need estimate_b3 computed above
coverage_table <- estimate_b3[, .(freq_covered = sum(beta_3 > lower &
                                 beta_3 < upper)/niter),
                            by = .(model, n2, sigma_1, rho, beta_1, beta_3)]

```

```{r coverage-plot}
coverage_table[, sim := factor(rep(1:8, 3))]
pal_okabe_ito_man <- pal_okabe_ito[c(5,7,3,6,1,2,4)]

pd <- position_dodge(0)
base <- 0.98
spacer <- 0.01
ggplot(data = coverage_table,
       aes(x = sim,
           y = freq_covered,
           color = model)) +
  geom_point(size = 3, position = pd) +
  geom_line(aes(group = model), position = pd) +
  
  annotate(geom = "text",
           label = c("n:"),
           x = 8.5,
           y = base + 2*spacer
  ) +
  annotate(geom = "text",
           label = coverage_table[1:8, n2],
           x = 1:8,
           y = base + 2*spacer
  ) +
  
  annotate(geom = "text",
           label = c("rho:"),
           x = 8.5,
           y = base + 1*spacer
  ) +
  annotate(geom = "text",
           label = coverage_table[1:8, rho],
           x = 1:8,
           y = base + 1*spacer
  ) +
 
  annotate(geom = "text",
           label = c("beta_3:"),
           x = 8.5,
           y = base + 0*spacer
  ) +
  annotate(geom = "text",
           label = coverage_table[1:8, beta_3],
           x = 1:8,
           y = base + 0*spacer
  ) +
 
  #coord_cartesian(xlim(c(0, 9))) +
  scale_color_manual(values = pal_okabe_ito_man,
                     name = NULL) +

  theme_pubr() +
  NULL

```

## Power

```{r}
# p-values
power_p <- melt(prob_power_table,
                id.vars = c("n2", "sigma_1", "beta_1", "beta_3", "rho"),
                measure.vars = c("lm_cov.p", "lm_change.p", "lda.p", "rmanova.p"),
                variable.name = "model",
                value.name = "p_value")
# N-P power
power_table <- power_p[, .(freq_lt_05 = sum(p_value < 0.05)/niter),
                       by = .(model, n2, sigma_1, rho, beta_1, beta_3)]

# Fisher sensitivity
quantile_list <- c(0.1, 0.25, 0.5, 0.75, 0.9)

lm_cov_p <- data.table(Model = "lm_cov",
                     t(quantile(power_p[model == "lm_cov.p", p_value],
                                quantile_list)))
lm_change_p <- data.table(Model = "lm_cov",
                     t(quantile(power_p[model == "lm_change.p", p_value],
                                quantile_list)))
lda_p <- data.table(Model = "lda",
                  t(quantile(power_p[model == "lda.p", p_value],
                             quantile_list)))
rmanova_p <- data.table(Model = "rmanova",
                      t(quantile(power_p[model == "rmanova.p", p_value],
                                 quantile_list)))

sensitivity_table <- rbind(lm_cov_p, lda_p, rmanova_p)
```

```{r power-plot}
power_table[, sim := factor(rep(1:8, 4))]
pal_okabe_ito_man <- pal_okabe_ito[c(5,7,3,6,1,2,4)]
ggplot(data = power_table,
       aes(x = sim,
           y = freq_lt_05,
           color = model)) +
  geom_point(size = 3) +
  geom_line(aes(group = model)) +
  
  annotate(geom = "text",
           label = c("n:"),
           x = 8.5,
           y = 1.1
  ) +
  annotate(geom = "text",
           label = power_table[1:8, n2],
           x = 1:8,
           y = 1.1
  ) +
  
  annotate(geom = "text",
           label = c("rho:"),
           x = 8.5,
           y = 1
  ) +
  annotate(geom = "text",
           label = power_table[1:8, rho],
           x = 1:8,
           y = 1
  ) +
 
  annotate(geom = "text",
           label = c("beta_3:"),
           x = 8.5,
           y = 0.9
  ) +
  annotate(geom = "text",
           label = power_table[1:8, beta_3],
           x = 1:8,
           y = 0.9
  ) +
 
  #coord_cartesian(xlim(c(0, 9))) +
  scale_color_manual(values = pal_okabe_ito_man,
                     name = NULL) +

  theme_pubr() +
  NULL

```

