---
title: Paired t-test as a special case of linear model and hierarchical (linear mixed)
  model
author: Jeff Walker
date: '2019-06-25'
slug: paired-t-test-as-a-special-case-of-linear-model-and-hierarchical-linear-mixed-model
categories:
  - stats 101
tags:
  - linear mixed model
  - microbiome
  - reanalysis
keywords:
  - tech
---

**Update** -- Fig. 2A is an analysis of the maximum endurance over three trials. [This has consequences.](../analyze-the-mean-or-median-and-not-the-max-response/)

Figure 2A of the paper [Meta-omics analysis of elite athletes identifies a performance-enhancing microbe that functions via lactate metabolism](https://www.nature.com/articles/s41591-019-0485-4) uses a paired t-test to compare endurance performance in mice treated with a control microbe (*Lactobacillus bulgaricus*) and a test microbe (*Veillonella atypica*) in a cross-over design (so each mouse was treated with both bacteria). The data are in the "Supplementary Tables" Excel file, which includes the raw data for the main paper.

### Setup

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggpubr)
library(readxl)
library(data.table)
library(lmerTest)
library(emmeans)

bookdown_it <- TRUE
if(bookdown_it==TRUE){
  data_path <- "../data"
  source("../../../R/clean_labels.R")
}else{
  data_path <- "../content/data"
  source("../../R/clean_labels.R")
}
```

### Import
```{r, message=FALSE}
folder <- "Data from Meta-omics analysis of elite athletes identifies a performance-enhancing microbe that functions via lactate metabolism"
file_i <- "41591_2019_485_MOESM2_ESM.xlsx"
file_path <- paste(data_path, folder, file_i, sep="/")
sheet_i <- "ST3 Veillonella run times"
range_vec <- c("a5:d21", "f5:i21", "a24:d40", "f24:i40")
fig2a <- data.table(NULL)
poop <- c("Lb", "Va")
for(week_i in 1:4){
  range_i <- range_vec[week_i]
  if(week_i %in% c(1, 3)){
    treatment <- rep(c(poop[1], poop[2]), each=8)
  }else{
    treatment <- rep(c(poop[2], poop[1]), each=8)
  }
  fig2a <- rbind(fig2a, data.table(week=week_i, treatment=treatment, data.table(
    read_excel(file_path, sheet=sheet_i, range=range_i))))
}

# clean column names
setnames(fig2a, old="...1", new="ID")
setnames(fig2a, old=colnames(fig2a), new=clean_label(colnames(fig2a)))

# wide to long
fig2a <- melt(fig2a, id.vars=c("week", "ID", "treatment"),
              variable.name="day",
              value.name="time")

```

Because of the cross-over design, the ID is not in order within each level of treatment. Make a wide data table with new treatment columns matched by ID.

```{r}
# get max for each week x ID x treatment combo
fig2a.max <- fig2a[, .(time_max=max(time)), by=.(week, ID, treatment)]

# match the two treatments applied to each ID
va <- fig2a.max[treatment=="Va", ]
lb <- fig2a.max[treatment=="Lb", ]
fig2a_wide <- merge(va, lb, by="ID")

```

### Comparison
#### paired t-test
```{r}
# paired t
res.t <- t.test(fig2a_wide[, time_max.x], fig2a_wide[, time_max.y], paired=TRUE)
res.t
```

#### linear model
```{r}
# as a linear model
y <- fig2a_wide[, time_max.x] - fig2a_wide[, time_max.y] # Va - Lb
res.lm <- lm(y ~ 1)
coef(summary(res.lm))

```

#### hierarchical (linear mixed) model with random intercept
```{r}
# as multi-level model with random intercept
res.lmer <- lmer(time_max ~ treatment + (1|ID), data=fig2a.max)
coef(summary(res.lmer))

```

