---
title: Can a linear model reproduce a Welch t-test?
author: Jeff Walker
date: '2019-10-27'
slug: can-a-linear-model-reproduce-a-welch-t-test
categories:
  - stats 101
tags:
  - p-values
  - linear model
keywords:
  - tech
---

This doodle was motivated Jake Westfall's answer to a [Cross-Validated question](https://stats.stackexchange.com/questions/142685/equivalent-to-welchs-t-test-in-gls-framework).

The short answer is yes but most R scripts that I've found on the web are unsatisfying because only the *t*-value reproduces, not the df and *p*-value. Jake notes the reason for this in his answer on Cross-Validated.

To get the adjusted df, and the *p*-value associated with this, one can use the emmeans package by Russell Lenth, [as he notes here](https://cran.rstudio.com/web/packages/emmeans/vignettes/FAQs.html).

# Setup
## Libraries

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(emmeans)
library(nlme)
```

## Functions

```{r functions}
reglance <- function(x){
  if(class(x)=="htest"){
    t <- x$statistic
    df <- x$parameter
    p <- x$p.value
  }
   if(class(x)=="lm"){
    df <- x$df.residual
    t <- coef(summary(x))["groupB", "t value"]
    p <- coef(summary(x))["groupB", "Pr(>|t|)"]
  }
 if(class(x)=="gls"){
    N <- summary(x)$dims$N
    k <- 2
    df <- N - k
    t <- summary(x)$tTable["groupB", "t-value"]
    p <- summary(x)$tTable["groupB", "p-value"]
  }
  if(class(x)=="emmGrid"){
    df <- summary(x)[1, "df"]
    t <- summary(x)[1, "t.ratio"]
    p <- summary(x)[1, "p.value"]
  }
  return(c(df=df, t=t, p=p))
}
```

# Motivating fake data
This is Jake's script, except that I've assigned the model objects and use these to make a table. I've also included the emmeans functions to generate the *t*-test using the gls model output.

```{r}
set.seed(497203)
dat <- data.frame(group=rep.int(c("A","B"), c(10,20)),
  y = rnorm(30, mean=rep.int(c(0,1), c(10,20)), sd=rep.int(c(1,2),c(10,20))))

# the t-statistic assuming equal variances
t.student <- t.test(y ~ group, data = dat, var.equal = TRUE)
m1 <- lm(y ~ group, data = dat)

# the t-statistic not assuming equal variances
t.welch <- t.test(y ~ group, data = dat, var.equal = FALSE)
m2 <- gls(y ~ group, data = dat, 
          weights=varIdent(form = ~ 1 | group))

m2.emm <- contrast(emmeans(m2, specs="group"))

t_table <- data.table(rbind(
  c("Student t", reglance(t.student)),
  c("lm", reglance(m1)),
  c("Welch t", reglance(t.welch)),
  c("gls", reglance(m2)),
  c("emmeans", reglance(m2.emm))
))
colnames(t_table) <- c("method", "df", "t", "p")
knitr::kable(t_table)
```

The values aren't rounded because I want to see how close the gls and Welch *t* are. As Jake notes, the GLS model doesn't use the Satterthwaite df, so only the *t* reproduces. That said, *in this example*, the non-adjusted and Satterthwaite df are very close. Consequently, *in this example*, the difference in p-values are trivial with respect to how we might interpret a result. The emmeans computation of the df and p-value are not equal but are satisfyingly close to those of the Welch *t*, at least for me, and at least *in this example*.

# Fake data with small n

Here I repeat Jake's example but with a smaller sample and I reverse which sample is associated with the larger variance.

```{r}
set.seed(497203)
n1 <- 8
n2 <- 4
dat <- data.frame(group=rep.int(c("A","B"), c(n1,n2)),
  y = rnorm(n1+n2, mean=rep.int(c(0,1), c(n1,n2)), sd=rep.int(c(1,2),c(n1,n2))))

# the t-statistic assuming equal variances
t.student <- t.test(y ~ group, data = dat, var.equal = TRUE)
m1 <- lm(y ~ group, data = dat)

# the t-statistic not assuming equal variances
t.welch <- t.test(y ~ group, data = dat, var.equal = FALSE)
m2 <- gls(y ~ group, data = dat, 
          weights=varIdent(form = ~ 1 | group))

m2.emm <- contrast(emmeans(m2, specs="group"))

t_table <- data.table(rbind(
  c("Student t", reglance(t.student)),
  c("lm", reglance(m1)),
  c("Welch t", reglance(t.welch)),
  c("gls", reglance(m2)),
  c("emmeans", reglance(m2.emm))
))
colnames(t_table) <- c("method", "df", "t", "p")
knitr::kable(t_table)
```

In this example, the lack of the Satterthwaite adjustment of the df matters. The emmeans computation of the df and *p*-value is very satisfying.