---
title: Paired line plots
author: Jeff Walker
date: '2019-01-22'
slug: paired-line-plots
categories:
  - ggplot
tags:
  - fake data
  - repeated measures
---

ggplot scripts to draw figures like those in the Dynamic Ecology post [Paired line plots (a.k.a. “reaction norms”) to visualize Likert data](https://dynamicecology.wordpress.com/2019/01/21/paired-line-plots-a-k-a-reaction-norms-to-visualize-likert-data/#comments)

```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggpubr)
library(data.table)
```

```{r fake data}
set.seed(3)
n <- 40
self <- rbinom(n, 5, 0.25) + 1
others <- self + rbinom(n, 3, 0.5)
fd <- data.table(id=factor(rep(1:n, 2)),
                 who=factor(rep(c("self", "others"), each=n)),
                 stigma <- c(self, others))
```

The students are identified by the column "id". The students two responses (self and others) are joined by a line using the geom_line() layer. geom_line knows who to join with the "group=id" statement the aes function in line 1. The alpha setting draws transparent lines (alpha=1 is opaque) so the more lines that are superimposed the darker the line. The lines are slightly jittered in the vertical direction so that overlapping lines appear wider.

```{r ggplot}
jitter_mag <- 0.03
gg <- ggplot(data=fd, aes(x=who, y=stigma, group=id)) +
  geom_line(size=1,
            alpha=0.3, 
            position=position_jitter(w=0, h=jitter_mag)) +
  ylab("Amount of stigma (1=low, 6=high)") +
  xlab("Responding about views of others or self") +
  theme_pubr() +
  NULL
gg
```

