---
title: Reanalyzing data from Human Gut Microbiota from Autism Spectrum Disorder Promote
  Behavioral Symptoms in Mice
author: Jeff Walker
date: '2019-06-03'
slug: reanalyzing-data-from-human-gut-microbiota-from-autism-spectrum-disorder-promote-behavioral-symptoms-in-mice
categories:
  - reproducibility
tags:
  - linear mixed model
  - microbiome
---

**Update** - This post has been updated

A very skeletal analysis of

Sharon, G., Cruz, N.J., Kang, D.W., Gandal, M.J., Wang, B., Kim, Y.M., Zink, E.M., Casey, C.P., Taylor, B.C., Lane, C.J. and Bramer, L.M., 2019. Human Gut Microbiota from Autism Spectrum Disorder Promote Behavioral Symptoms in Mice. Cell, 177(6), pp.1600-1618.

[which got some attention on pubpeer](https://pubpeer.com/publications/B521D325772244D8F656F1ED193ACA#).

Commenters are questioning the result of Fig1G. It is very hard to infer a p-value from plots like these, where the data are multi-level, regardless of if means and some kind of error bar is presented. A much better plot for inferring differences is an effects plot with the CI of the effect. That said, I'll try to reproduce the resulting p-value.

# Caveats
Failure to reproduce or partial reproducibility might be an error in my coding, or my misunderstanding of the author's methods, or my lack of knowledge of statistics generally or the R functions that I use more specifically.

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(nlme)
library(lmerTest)
library(car)
library(emmeans)
library(lmtest)
library(data.table)
```

```{r}
data_path <- "../data/Data from Human Gut Microbiota from Autism Spectrum Disorder Promote Behavioral Symptoms in Mice/Fig1"
#data_path <- "data/Data from Human Gut Microbiota from Autism Spectrum Disorder Promote Behavioral Symptoms in Mice/Fig1"
```

```{r}
fn <- "Fig1EFGH_subset8.csv"
file_path <- paste(data_path, fn, sep="/")
mouse <- fread(file_path)
mouse[, Gender:=factor(Gender)]
mouse[, ASD_diagnosis:=factor(ASD_diagnosis, c("NT", "ASD"))]
mouse[, MouseID:=factor(MouseID)]
mouse[, Round:=factor(Round)]
```

# Figure 1G

## Estimate of marginal means and Cohen's d

One could reasonably estimate Cohen's d several ways given the 2 x 2 design. Here are two

```{r}
m1 <- lmer(OFT_Distance ~ ASD_diagnosis + (1|Donor), data=mouse)
(m1.emm <- emmeans(m1, specs=c("ASD_diagnosis")))
m2 <- lmer(OFT_Distance ~ Gender + ASD_diagnosis + (1|Donor), data=mouse)
(m2.emm <- emmeans(m2, specs=c("ASD_diagnosis")))

```

The effect, standardized to cohen's D using residual standard deviation from model: (NT - ASD)/sigma 

```{r}
coef(summary(m1))["ASD_diagnosisASD", "Estimate"]/sigma(m1)

summary(contrast(m2.emm, method="revpairwise"))[1, "estimate"]/sigma(m2)


```

The reported cohen's D for the distance traveled is doTD-oASD = -0.58

## Fig 1G p-value
### Method no. 1 (as stated in caption to Fig 1)
The figure caption gives the method as "Hypothesis testing for differences of the means were done by a mixed effects analysis using donor diagnosis and mouse sex as fixed effects and donor ID as a random effect. p values were derived from a chi-square test". This would be a Likelihood Ratio Test (LRT). The LRT requires that the model be fit by maximum likelhood for the statistic to have any meaning. The default fit in R is REML.

The LRT of the model as specified in the caption of Fig1G
```{r}

# correct LRT
m1 <- lmer(OFT_Distance ~ Gender + ASD_diagnosis + (1|Donor), REML=FALSE, data=mouse)
m2 <- lmer(OFT_Distance ~ Gender + (1|Donor), REML=FALSE, data=mouse)
lrtest(m2, m1)

```

The p-value is .2208

The LRT of the model as specified but using the default lmer settings, which is fit by REML, which results in a meaningless chisq statistic

```{r}

m1 <- lmer(OFT_Distance ~ Gender + ASD_diagnosis + (1|Donor), data=mouse)
m2 <- lmer(OFT_Distance ~ Gender + (1|Donor), data=mouse)
lrtest(m2, m1)
```

the p-value is 0.00013. Is this the reported p-value? I don't know. A problem is that the methods section suggests a different model was used. This leads to:

## method no. 2 (as stated in methods section)

"Statistical analysis for behavioral outcomes in fecal transplanted offspring. Comparison of behavioral outcomes between TD Controls and ASD donors were tested using longitudinal linear mixed effects analyses, with test cycles and donors treated as repeated factors. Analyses were performed in SPSS (v 24); a priori alpha = 0.05. All outcomes were tested for normality and transformed as required. Diagonal covariance matrices were used so that intra-cycle and intra-donor correlations were accounted for in the modeling. The donor type (TD versus ASD) was the primary fixed effect measured, and mouse sex was an a priori covariate."

(I think) it would be unlikely to get the wrong LRT using SPSS. Also the specified model differs from that in the caption. Was the response transformed (note it's not the reponse that needs to be "normal" but the residuals from the linear model (or equivalently the conditional response))

correct LRT
```{r}
# correct LRT
m1 <- lmer(OFT_Distance ~ Gender + ASD_diagnosis + (1|Cycle) + (1|Donor), REML=FALSE, data=mouse)
m2 <- lmer(OFT_Distance ~ Gender + (1|Cycle) + (1|Donor), REML=FALSE, data=mouse)
lrtest(m2, m1)
```

Hmm. Need to explore the error.

incorrect LRT

```{r}
# correct LRT
m1 <- lmer(OFT_Distance ~ Gender + ASD_diagnosis + (1|Cycle) + (1|Donor), data=mouse)
m2 <- lmer(OFT_Distance ~ Gender + (1|Cycle) + (1|Donor), data=mouse)
lrtest(m2, m1)
```

Basically the same results

# Update

1. [Thomas Lumley also explored statistical models based on the author's two descriptions](https://notstatschat.rbind.io/2019/06/16/analysing-the-mouse-autism-data/). The first set are similar to mine above, but using Satterthwaite or Kenward-Rogers ddf on F-test instead of a LRT. I punted on the "diagonal covariance matrices" of method 2, but Lumley ran with it and, with a bit of SPSS manual description, figured out that the result can be reproduced using a GLS that weights the error variances using levels of Donor. This isn't a mixed model with Donor as a random effect, as stated in the methods. Most importantly, since donor isn't a random effect, the model uses complete pooling, which ignores the correlated error arising from the clustered design (mice recieving microbes from the same donor). The consequence is a super liberal denominator df for the test, and overly optimistic p-values.

```{r}

type3 <- list(Gender=contr.sum, ASD_diagnosis=contr.sum) # change the contrasts coding for the model matrix
m2a <- lm(OFT_Distance ~ Gender*ASD_diagnosis, data=mouse, contrasts=type3)
Anova(m2a, type="3")
# contrasts statement doesn't work with gls so change variable in data.table
mouse[, Gender.sum := C(Gender, contr.sum)]
mouse[, ASD_diagnosis.sum := C(ASD_diagnosis, contr.sum)]
m2b <- gls(OFT_Distance ~ Gender.sum*ASD_diagnosis.sum, weights=varIdent(form=~1|Donor), data=mouse)
anova(m2b, type="marginal")

```


## some checks
### What is MouseID?

In the pubpeer response, Sharon states "specify linear mixed effects modeling of individual mice, nested within donor, and round of testing". I assume individual mice is referenced by MouseID. But...

Number of Mice
```{r}
length(levels(mouse$MouseID))
```

Explore a single mouse
```{r}
mouse[MouseID==1, 1:10]
# mouse ID=1 is both female and male? So MouseID cannot be the mouse tested behaviorally
```

Hmm. MouseID=1 is both male and female? Is this true of some/all MouseID?

```{r}
# check number of mouseID:Gender combinations
unique(paste(mouse$MouseID, mouse$Gender,sep=":"))

```

So 75 "individual mice" (identified by MouseID) but 106 combinations of MouseID:Gender. So I'm not sure what MouseID identifies -- maybe litter (see below). Certainly not an individual mouse that were tested.

### What is a cycle v round?
Again, in the pubpeer response, Sharon states "specify linear mixed effects modeling of individual mice, nested within donor, and round of testing

```{r}
dim(mouse)
length(unique(paste(mouse$MouseID, mouse$Round,sep=":")))
```

206 data points (consistant with results) but 204 combinations of MouseID and Round. Which have > 1
```{r}
mouse[, ID_x_Round:=paste(MouseID, Round, sep="_x_")]
dt <- mouse[, .(N=.N), by=ID_x_Round]
dt[N>1,] # 35 x 1 and 46 x 1
mouse[MouseID=="35", ]
mouse[MouseID=="46", ]

```

In addition to two 35 x 1 (mouse ID x Round), this same MouseID is used for mice of different sex and different donors. Does MouseID=litter?
