---
title: Melting a list of columns
author: Jeff Walker
date: '2020-04-27'
slug: melting-a-list-of-columns
categories:
  - data wrangling
tags:
  - data.table
  - melt
keywords:
  - tech
---

[An answer to this tweet](https://twitter.com/lpsatchell/status/1254763204061716484)
"Are there any #Rstats tidy expeRts who'd be interested in improving the efficiency of this code that gathers multiple variables from wide to long?

This works but it's not pretty. There must be a prettier way..."


Wide data frame has three time points where participants answer two questions on two topics.

# create data from original code

```{r}
#Simmed data
Time1.Topic1.Question1 <- rnorm(500)
data <- data.frame(Time1.Topic1.Question1)
data$Time1.TOpic1.Question2 <- rnorm(500)
data$Time1.Topic2.Question1 <- rnorm(500)
data$Time1.Topic2.Question2 <- rnorm(500)
data$Time2.Topic1.Question1 <- rnorm(500)
data$Time2.Topic1.Question2 <- rnorm(500)
data$Time2.Topic2.Question1 <- rnorm(500)
data$Time2.Topic2.Question2 <- rnorm(500)
data$Time3.Topic1.Question1 <- rnorm(500)
data$Time3.Topic1.Question2 <- rnorm(500)
data$Time3.Topic2.Question1 <- rnorm(500)
data$Time3.Topic2.Question2 <- rnorm(500)
data <- tibble::rowid_to_column(data, "Id")
data$Condition <- sample(x = 1:5, size=500, replace=TRUE)

#Starting data
# View(data)
```

# My code to melt on a list of columns

My code is a data.table solution. Melt is a beutiful function and data.table made it better. I follow the original code (below) and name the columns using "T" for topic (e.g. T1Q1) but the future me might think this stands for "Time" and not "Topic". I'd probably name the columns something like "topic_1_Q_1"

```{r}
library(data.table)
dt <- data.table(data)
setnames(dt, "Time1.TOpic1.Question2", "Time1.Topic1.Question2")
T1Q1 <- colnames(dt)[grep("Topic1.Question1", colnames(dt))]
T1Q2 <- colnames(dt)[grep("Topic1.Question2", colnames(dt))]
T2Q1 <- colnames(dt)[grep("Topic2.Question1", colnames(dt))]
T2Q2 <- colnames(dt)[grep("Topic2.Question2", colnames(dt))]

dt_long <- melt(dt, 
                id.vars = c("Id"),
                measure.vars = list(T1Q1, T1Q2, T2Q1, T2Q2),
                variable.name <- "Time",
                value.name = c("T1Q1", "T1Q2", "T2Q1", "T2Q2"))
# View(dt_long)
```

## notes on my code

The grep statements are key. `grep("Topic1.Question1", colnames(dt))` finds the columns in which the column name includes the text in the quotes. Then embedding this within `colnames(dt)[]` returns the names of the indexed columns. For example, `grep("Topic1.Question1", colnames(dt))` returns the vector

```{r echo=FALSE}
grep("Topic1.Question1", colnames(dt))
```

and `colnames(dt)[grep("Topic1.Question1", colnames(dt))]` returns the vector

```{r echo=FALSE}
colnames(dt)[grep("Topic1.Question1", colnames(dt))]
```

# Original code
```{r}

#Needs to become a long data frame for time, but independent columnsfor topic and question.
#The clunky way to do this is my solution below. Is there a more efficient solution?
library(dplyr)
library(tidyr)

#Simmed data
Time1.Topic1.Question1 <- rnorm(500)
data <- data.frame(Time1.Topic1.Question1)
data$Time1.TOpic1.Question2 <- rnorm(500)
data$Time1.Topic2.Question1 <- rnorm(500)
data$Time1.Topic2.Question2 <- rnorm(500)
data$Time2.Topic1.Question1 <- rnorm(500)
data$Time2.Topic1.Question2 <- rnorm(500)
data$Time2.Topic2.Question1 <- rnorm(500)
data$Time2.Topic2.Question2 <- rnorm(500)
data$Time3.Topic1.Question1 <- rnorm(500)
data$Time3.Topic1.Question2 <- rnorm(500)
data$Time3.Topic2.Question1 <- rnorm(500)
data$Time3.Topic2.Question2 <- rnorm(500)
data <- tibble::rowid_to_column(data, "Id")
data$Condition <- sample(x = 1:5, size=500, replace=TRUE)

#Starting data
# View(data)

#Pull out sections and long-ify Topics
data %>% select(Id, ends_with("Topic1.Question1")) -> T1.Q1.data 
T1.Q1.data <- gather(T1.Q1.data, Time, T1Q1, 2:4, factor_key = TRUE)

data %>% select(Id, ends_with("Topic1.Question2")) -> T1.Q2.data 
T1.Q2.data <- gather(T1.Q2.data, Time, T1Q2, 2:4, factor_key = TRUE)

data %>% select(Id, ends_with("Topic2.Question1")) -> T2.Q1.data 
T2.Q1.data <- gather(T2.Q1.data, Time, T2Q1, 2:4, factor_key = TRUE)

data %>% select(Id, ends_with("Topic2.Question2")) -> T2.Q2.data 
T2.Q2.data <- gather(T2.Q2.data, Time, T2Q2, 2:4, factor_key = TRUE)

#Create a time coded variable that doesn't contain mess of column title names 
data %>% select(Id) -> time.data
time.data$Baseline <- 0
time.data$TimeOne <- 1
time.data$Timetwo <- 2
time.data <- gather(time.data, Time, Timepoint, 2:4, factor_key = TRUE)

#Drop uneededs
time.data %>% select(-(Time)) -> time.data
T1.Q1.data %>% select(-(Time)) -> T1.Q1.data
T1.Q2.data %>% select(-(Time)) -> T1.Q2.data
T2.Q1.data %>% select(-(Time)) -> T2.Q1.data
T2.Q2.data %>% select(-(Time)) -> T2.Q2.data

#Merge
data %>% select(Id, Condition) -> idcond.data 
longdata <- merge(idcond.data,time.data,by="Id")
longdata <- merge(longdata,T1.Q1.data,by="Id")
longdata <- merge(longdata,T1.Q2.data,by="Id")
longdata <- merge(longdata,T2.Q1.data,by="Id")
longdata <- merge(longdata,T2.Q2.data,by="Id")

#View
# View(longdata)
```
