---
title: Analyzing longitudinal data -- a simple pre-post design
author: Jeff Walker
date: '2020-03-19'
slug: analyzing-longitudinal-data-a-simple-pre-post-design
categories:
  - stats 101
tags:
  - ancova
  - fake data
  - linear mixed model
  - NHST
  - power
  - pre-post
  - clda
keywords:
  - tech
---


A skeletal response to a twitter question:

"ANOVA (time point x group) or ANCOVA (group with time point as a covariate) for intervention designs? Discuss."

follow-up "Only 2 time points in this case  (pre- and post-intervention), and would wanna basically answer the question of whether out of the 3 intervention groups, some improve on measure X more than others after the intervention"

Here I compare five methods using fake pre-post data, including

1. lm-cov. A linear model with baseline as a covariate. This is "ANCOVA" in the ANOVA-world.
2. clda. constrained longitudinal data analysis (cLDA). A linear mixed model in which the intercept is constrained to be equal (no treatment effect at time 0).
3. rmanova. Repeated measures ANOVA.
4. lmm. A linear mixed model version of rmanova. A difference is the rmanova will use a sphericity correction.
5. two-way. Linear model with Treatment and Time and their interaction as fixed factors. Equivalent to a Two-way fixed effects ANOVA in ANOVA-world.

I use a simulation to compute Type I error rate and Power. This is far from a comprehensive simulation. Like I said, its skeletal. 

TL;DR. At the risk of concluding anything from a very skeletal simulation:

* lm-cov has best performance in since of combining control of Type I error and relative power
* smaller SE in clda comes at a cost of relatively high Type I error but increased powers. Do with this what you will.
* rmanova has smaller power than lm-cov

Not included: I didn't save the results necessary to plot type I error conditional on baseline difference, which is well controlled by lm-cov (and clda) but not by lmm or rmanova.

# libraries
```{r libraries, message=FALSE, warning=FALSE}
library(data.table)
library(nlme)
library(emmeans)
library(afex)
library(ggpubr)
library(ggsci)
library(cowplot)
```

# simulation functions
## pre-post fake data generator 

My general algorithm for creating pre-post data with a single post time point differs from how I create pre-post data with multiplee post time points.

```{r pre-post-FDG}

generate_pre_post <- function(
  n = 6, # if vector then sample per group
  sigma_0 = 1, # error at time 0
  sigma_1 = 1, # error at time 1
  rho = 0.6, # correlation between time 0 and 1
  beta_0 = 10, # control mean at time 0
  beta_1 = 1, # treatment effect at time 1
  beta_2 = 5 # time effect at time 1
){
  
  if(length(n) == 1){n <- rep(n,2)} # set n = n for all groups
  N <- sum(n)
  
  fd <- data.table(
    treatment = factor(rep(c("cn", "tr"), n))
  )
  X <- model.matrix(~treatment, data=fd)
  
  z <- rnorm(N) # z is the common factor giving correlation at time 0 and 1. This
  # is easier for me than making glucose_1 = glucose_0 + treatment + noise
  fd[, glucose_0 := beta_0 + sqrt(rho)*z + sqrt(1 - rho)*rnorm(N, sd=sigma_0)]
  fd[, glucose_1 := beta_0 + sqrt(rho)*z*sigma_1 + sqrt(1 - rho)*rnorm(N, sd=sigma_1) +
       X[,2]*beta_1 + # add treatment effect
       beta_2 # add time effect
     ]
  fd[, change := glucose_1 - glucose_0] # change score
  fd[, id := factor(1:.N)]
  return(fd)
}

```

## test FDG

```{r test pre-post-FDG}
fd <- generate_pre_post(
  n = c(10^5, 10^5),
  sigma_0 = 1,
  sigma_1 = 2,
  rho = 0.6, 
  beta_0 = 10, 
  beta_1 = 1 
)

m1 <- lm(glucose_1 ~ treatment, data = fd)
fd[, glucose_1_cond := coef(m1)[1] + residuals(m1)]
coef(summary(m1))
cov(fd[, .SD, .SDcols=c("glucose_0", "glucose_1_cond")])
cor(fd[, .SD, .SDcols=c("glucose_0", "glucose_1_cond")])
coef(summary(lm(glucose_1 ~ glucose_0 + treatment, data=fd)))

```


## fake data wide to long

```{r wide-to-long}
wide_to_long <- function(
  dt, #data.table,
  measure_prefix = "glucose"
){
# measure _prefix is the prefix of the measurement columns that will be melted. A "_" between the prefix and a time value is assumed. So with measure vars like glucose_0, glucose_15, glucose_20, the measure prefix is "glucose"
  dt <- data.table(dt)
  dt_long <- melt(dt, 
                  id.vars=c("treatment", "id"), 
                  measure.vars=patterns(paste0(measure_prefix, "_")), 
                  variable.name = "time",
                  value.name = measure_prefix)
  return(dt_long)
}
```

# Fit model functions
## two-way anova

```{r two-way anova}
two_way <- function(dt_long){
  fit <- lm(glucose ~ time*treatment, data=dt_long)
  return(fit)
}
```

## repeated measures anova

```{r repeated-measures-anova, message=FALSE}
rmanova <- function(dt_long){
  fit <- aov_4(glucose ~ time*treatment + (time|id),
               data=dt_long)
  return(fit)
}
```

## linear model with baseline as covariate

```{r lm-cov}
lm_cov <- function(dt){
  fit <- lm(glucose_1 ~ glucose_0 + treatment, data = dt)
  return(fit)
}
```

## linear mixed model

```{r lmm}
lmm <- function(dt_long){
  fit <- lme(glucose ~ time*treatment,
             random = ~1|id,
             data = dt_long,
             weights = varIdent(form= ~ 1 | time),
             correlation= corSymm(form=~ 1 | id))
  return(fit)
}
```

## cLDA functions

Two equivalent functions for CLDA

### clda-1

```{r clda-1}
clda1 <- function(dt_long){
  design <- model.matrix( ~ time + treatment:time, data = dt_long)
  # remove intercept column and effect of tr at time 0
  X <- design[, -c(1, which(colnames(design) == "timeglucose_0:treatmenttr"))]
  colnames(X)[2] <- "timeglucose_1_treatmenttr"
  dt_long_x <- cbind(dt_long, X)
  form <- formula(paste0("glucose ~ ", paste(colnames(X), collapse = " + ")))
  fit <- gls(form,
            data = dt_long_x,
            weights = varIdent(form= ~ 1 | time),
            correlation= corSymm(form=~ 1| id))
  return(fit)
}
```

### clda-2

```{r clda-2}
clda2 <- function(dt_long){
  dt_clda <- copy(dt_long)
  dt_clda[, time.treatment := ifelse(time != "glucose_0" & treatment=="tr", paste0(time, ":tr"), "cont")]
  dt_clda[, time.treatment := factor(time.treatment, c("cont","glucose_1:tr"))]
  fit <- gls(glucose ~ time + time.treatment,
            data = dt_clda,
            weights = varIdent(form= ~ 1 | time),
            correlation= corSymm(form=~ 1| id))
  return(fit)
}

```

## test functions

Some equivalents

1. cLDA is equivalent estimate to the linear model with baseline covariate but has a smaller standard error.
2. the two-way (fixed) ANOVA and lmm have equivalent estimates but SE is larger
3. the lmm and rmanova have the same p-value (assuming sphericity)

```{r test, message=FALSE}
fd <- generate_pre_post(
  n = 6,
  sigma_0 = 1,
  sigma_1 = 1,
  rho = 0.6, 
  beta_0 = 10, 
  beta_1 = 1 
)

fd_long <- wide_to_long(fd,
                        measure_prefix = "glucose"
                        )

m1 <- lm_cov(fd)
m2 <- clda1(fd_long)
m3 <- clda2(fd_long)
m4 <- lmm(fd_long)
m5 <- two_way(fd_long)
m6 <- rmanova(fd_long)
```

linear model with baseline covariate
```{r test-results-lm-cov}
coef(summary(m1))
```

cLDA (both R methods)
```{r test-results-clda}
coef(summary(m2))
coef(summary(m3))
```

two-way (fixed effect) linear model
```{r test-results-2-way-fixed}
coef(summary(m5))
```

linear mixed model
```{r test-results-lmm}
coef(summary(m4))
```


repeated measures ANOVA
```{r test-results-rmanova}
summary(m6)
```

# simulations
## single post time point ("pre-post")

```{r simulation-single-post-time-point, message=FALSE}

simulate_pre_post <- function(
  niter = 1000, # number of iterations
  n = 6, # if vector then sample per group
  sigma_0 = 1, # error at time 0
  sigma_1 = 1, # error at time 1
  rho = 0.6, # correlation between time 0 and 1
  beta_0 = 10, # control mean at time 0
  beta_1 = 1, # treatment effect at time 1
  beta_2 = 5 # time effect at time 1
){
  
  if(length(n) == 1){n <- rep(n,2)} # set n = n for all groups
  method_list <- c("lm_cov", "clda", "lmm", "rmanova")
  prob_wide <- data.table(matrix(-9999, nrow=niter, ncol=length(method_list)))
  colnames(prob_wide) <- method_list
  for(iter in 1:niter){
    fd <- generate_pre_post(
      n = n,
      sigma_0 = sigma_0,
      sigma_1 = sigma_1,
      rho = rho, 
      beta_0 = beta_0, 
      beta_1 = beta_1 
    )
    
    fd_long <- wide_to_long(fd,
                            measure_prefix = "glucose"
    )
    
    m1 <- lm_cov(fd)
    m2 <- clda1(fd_long)
    m3 <- clda2(fd_long)
    m4 <- lmm(fd_long)
    m5 <- two_way(fd_long)
    m6 <- rmanova(fd_long)
    
    prob_wide[iter, lm_cov := coef(summary(m1))["treatmenttr", "Pr(>|t|)"]]
    prob_wide[iter, clda := coef(summary(m2))["timeglucose_1_treatmenttr", "p-value"]]
    prob_wide[iter, lmm := coef(summary(m4))["timeglucose_1:treatmenttr", "p-value"]]
    prob_wide[iter, rmanova := summary(m6)$univariate.tests["treatment:time", "Pr(>F)"]]
    
  }
  
  prob <- melt(prob_wide,
               id.vars = NULL,
               measure.vars = method_list,
               variable.name = "method",
               value.name = "p_value")
  return(prob)
}
```

## Summary statistics

```{r summary stats}
p_summary <- function(prob, niter){
  prob_summary <- prob[, .(freq_lt_05 = sum(p_value < 0.05)/max(niter)),
          by = .(method)]
  return(prob_summary)
}

```

## Plot: 

```{r plot-it}
plot_it <- function(prob){
  jco_pal <- pal_jco()(6)
  group_colors <- jco_pal[c(2,5,4,6)]
  group_shapes <- c(15, 16, 17,  18)
  
  pd <- position_dodge(0.8)
  gg1 <- ggplot(data = prob,
                aes(x=method, y=freq_lt_05, color=method, shape=method)) +
    geom_point(position = pd) +
    scale_color_manual(values = group_colors) +
    scale_shape_manual(values = group_shapes) +
    ylab("Frequency p < 0.05") +
    xlab("") +
    theme_pubr() +
    theme(legend.position="none") +
    guides(col = guide_legend(ncol = 5, byrow = FALSE)) +
    NULL
  return(gg1)
}
```

## Type I error

```{r type I, message=FALSE, eval=TRUE}
# parameters of simulation
  niter = 1000 # number of iterations
  n = 6 # if vector then sample per group
  sigma_0 = 1 # error at time 0
  sigma_1 = 1 # error at time 1
  rho = 0.6 # correlation between time 0 and 1
  beta_0 = 10 # control mean at time 0
  beta_1 = 0 # treatment effect at time 1
  beta_2 = 5 # time effect at time 1
  
  prob_type1 <- simulate_pre_post(niter = niter,
                                  n = n,
                                  sigma_0 = sigma_0,
                                  sigma_1 = sigma_1,
                                  rho = rho,
                                  beta_0 = beta_0,
                                  beta_1 = beta_1,
                                  beta_2 = beta_2)

prob_type1_sum <- p_summary(prob_type1, niter)
```

## Power

```{r Power sim, message=FALSE, eval=TRUE}
# parameters of simulation
  niter = 1000 # number of iterations
  n = 6 # if vector then sample per group
  sigma_0 = 1 # error at time 0
  sigma_1 = 1 # error at time 1
  rho = 0.6 # correlation between time 0 and 1
  beta_0 = 10 # control mean at time 0
  beta_1 = 1 # treatment effect at time 1
  beta_2 = 5 # time effect at time 1

  prob_power <- simulate_pre_post(niter = niter,
                              n = n,
                              sigma_0 = sigma_0,
                              sigma_1 = sigma_1,
                              rho = rho,
                              beta_0 = beta_0,
                              beta_1 = beta_1,
                              beta_2 = beta_2)
  
  prob_power_sum <- p_summary(prob_power, niter)
```

# Tables
## Type I error rates
```{r type I table, eval=TRUE}
table1 <- prob_type1_sum
colnames(table1)[2] <- "Type I error rate"
knitr::kable(table1)
```

## Power

```{r power table, eval=TRUE}
table2 <- prob_power_sum
colnames(table2)[2] <- "Power"
knitr::kable(table2, digits = 2)
```

# Plots
```{r plots}
  
gg_a <- plot_it(prob_type1_sum)
gg_b <- plot_it(prob_power_sum)
plot_grid(gg_a, gg_b, ncol=2)

```


